---
title: "Assignement 2"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
date: "2024-01-10"
---

## Introduction :

This assignment consists of performing data cleaning and manipulation, and then some statistical analysis. 

## Dataset:

The dataset is retrieved from [Rwanda DHS (Demographic and Health Survey)](https://dhsprogram.com/Data/) 2020. The type of dataset used here is Household member. You will get data in two files: main SPSS File and Map File (for descriptions). 

Your Assignments steps:

1. Read the dataset in R.

```{r echo=TRUE}
# required package
# install.packages("haven")
library(haven)

# import data
Dhs_data<-read_sav("C:/Users/RPHC5/Downloads/Assignment 2_files/DHS2020/RWPR81FL.SAV")

```

- Visualize, inspect and get familiar with the data
```{r echo=FALSE}
# your code here
dim(Dhs_data)
```


2. Select only few columns, important in this Assignments. They are the following: "HV001", "HV009", "HV010", "HV011", "HV014", "SHDISTRICT", "HV024", "HV025", "HV040", "HV227", "HV228", "HV270", "HV105", "HV106", "HML3", "HML4", "HML7",
  "HML10", "HML22", "HML32","HML33", "HML35"

```{r echo=FALSE}
# your code here
Dhs_data1 <- data.frame(Dhs_data$HV001, Dhs_data$HV009, Dhs_data$HV010, Dhs_data$HV011, Dhs_data$HV014, Dhs_data$SHDISTRICT, Dhs_data$HV024, Dhs_data$HV025, Dhs_data$HV040, Dhs_data$HV227, Dhs_data$HV228, Dhs_data$HV270, Dhs_data$HV105, Dhs_data$HV106, Dhs_data$HML3, Dhs_data$HML4, Dhs_data$HML7, Dhs_data$HML10, Dhs_data$HML22, Dhs_data$HML32, Dhs_data$HML33, Dhs_data$HML35)

```

  
3. Rename variables using the variable descriptions below. Give meaningful (short) name to the variables of your choice.

>* HV001= "Cluster number", Cl_NUMBER
>* HV009 = "Number of household members", N-HHD
>* HV010 = "Number of eligible women in household",
>* HV011 = "Number of eligible men in household",
>* HV014 = "Number of children 5 and under (de jure)",
>* SHDISTRICT = "District (geographic area)",
>* HV024 = "Region (provinces, corresponding values in a map file)",
>* HV025 = "Type of place of residence (rural versus urban)",
>* HV040 = "Cluster altitude in meters",
>* HV227 = "Presence of mosquito bed net for sleeping",
>* HV228 = "Number of children under 5 who slept under a mosquito bed net",
>* HV270 = "Wealth index combined (an index based on various household assets indicating socio-economic status)",
>* HV105 = "Age of household members",
>* HV106 = "Highest educational level attained by individuals",
>* HML3 = "Net observed by interviewer",
>* HML4 = "Months ago the net was obtained",
>* HML7 = "Brand of net",
>* HML10 = "Insecticide-Treated Net (ITN)",
>* HML22 = "Obtained net from campaign, antenatal, or immunization visit",
>* HML33= "Result of malaria measurement", 
>* HML32 = "Final result of malaria from blood smear test",
>* HML35 = "Result of malaria rapid test"

```{r echo=FALSE}
colnames(Dhs_data1) <- c("Cl_NUMBER", "N-HHD", "N_El_Women", "N_El_Men", "N_Child_5","District",
"Region", "Type_Residence", "Cl_Altitude_ M", "Pre_mosquito_net", "N_Child_U_5", "W_Index", "Age", "Educ_level","Net_observed", "Months_Ago", "Brand_net", "Ins_Treat_N", "Ob_net_Antenatal", "Blood_smear_test", "Malaria_measurement", "Malaria _rapid_test")
```


## Data cleaning

1. Inspect each variables, decode variable to its original unique variables. 
Example, Variable "HV024"(Region) has Unique values 1,2,3,4,5. Decode it to orginal Region Kigali, South, West, North, East
Use Map file to see the description of each values in data. 

```{r echo=FALSE}
# your code
#install.packages("tidyverse")
library("tidyverse")
all_columns <-c("Cl_NUMBER", "N-HHD", "N_El_Women", "N_El_Men", "N_Child_5","District",
"Region", "Type_Residence", "Cl_Altitude_ M", "Pre_mosquito_net", "N_Child_U_5", "W_Index", "Age", "Educ_level","Net_observed", "Months_Ago", "Brand_net", "Ins_Treat_N", "Ob_net_Antenatal", "Blood_smear_test", "Malaria_measurement", "Malaria _rapid_test")
Dhs_data1[all_columns] <- sapply(Dhs_data1[all_columns], as_factor)


```
(Work process in attached RMD file)


2. Handling Missing Values:

Determine columns with missing values. Devise the strategy to handle missing values: Deleting missing values, replacing missing values with mean or mode.
```{r echo=FALSE}
# Check for missing values in the entire dataset
missing_values <- colSums(is.na(Dhs_data1))

# View columns with missing values
missing_values[missing_values > 0]

mode <- function(x) names(sort(table(x), decreasing = TRUE))[1]
Dhs_data1$N_Child_U_5[is.na(Dhs_data1$N_Child_U_5)] <- mode(Dhs_data1$N_Child_U_5)

mode <- function(x)names(sort(table(x), decreasing = TRUE))[1] 
Dhs_data1$Mal_Res_Rapid[is.na(Dhs_data1$Mal_Res_Rapid)] <- mode(Dhs_data1$Mal_Res_Rapid)


mode<-function(x)(names(sort(table(x), decreasing = TRUE))[1])
Dhs_data1$Mal_Final[is.na(Dhs_data1$Mal_Final)] <- mode(Dhs_data1$Mal_Final)


mode<-function(x)(names(sort(table(x), decreasing = TRUE))[1])
Dhs_data1$Brand_net[is.na(Dhs_data1$Brand_net)] <- mode(Dhs_data1$Brand_net)


mode<-function(x)(names(sort(table(x), decreasing = TRUE))[1])
Dhs_data1$Net_observed[is.na(Dhs_data1$Net_observed)] <- mode(Dhs_data1$Net_observed)


mode<-function(x)names(sort(table(x), decreasing = TRUE))[1]
Dhs_data1$Ins_Treat_N[is.na(Dhs_data1$Ins_Treat_N)] <- mode(Dhs_data1$Ins_Treat_N)


mode<-function(x)(names(sort(table(x), decreasing = TRUE))[1])
Dhs_data1$Ob_net_Antenatal[is.na(Dhs_data1$Ob_net_Antenatal)] <- mode(Dhs_data1$Ob_net_Antenatal)


mode<-function(x)(names(sort(table(x), decreasing = TRUE))[1])
Dhs_data1$Mal_Res_Blood[is.na(Dhs_data1$Mal_Res_Blood)] <- mode(Dhs_data1$Mal_Res_Blood)


Dhs_data1$Net_Mon_Old_Original <- Dhs_data1$Net_Mon_Old  
Dhs_data1$Net_Mon_Old <- (as.character(Dhs_data1$Months_Ago)) 


mode<-function(x)(names(sort(table(x), decreasing = TRUE))[1])
Dhs_data1$Months_Ago[is.na(Dhs_data1$Months_Ago)] <- mode(Dhs_data1$Months_Ago)

mode<-function(x)(names(sort(table(x), decreasing = TRUE))[1])
Dhs_data1$Malaria_measurement[is.na(Dhs_data1$Malaria_measurement)] <- mode(Dhs_data1$Malaria_measurement)

mode<-function(x)(names(sort(table(x), decreasing = TRUE))[1])
Dhs_data1$Blood_smear_test[is.na(Dhs_data1$Blood_smear_test)] <- mode(Dhs_data1$Blood_smear_test)

mode<-function(x)(names(sort(table(x), decreasing = TRUE))[1])
Dhs_data1$`Malaria _rapid_test`[is.na(Dhs_data1$`Malaria _rapid_test`)] <- mode(Dhs_data1$`Malaria _rapid_test`)


# your code
```


3. Create new variables

>a. Create variable called "Old Mosquito" variable HML4 (Months ago the net was obtained). The created variable must binary with 1 when mosquito is more than 24 months old. 
>b. Create Variable "Average District altitude". Create this variable by averaging cluster altitude in each district. We have three variables HV001= "Cluster number", SHDISTRICT = "District (geographic area)" and HV040 = "Cluster altitude in meters". Filter out clusters in each district, do `mean` of cluster altitude in that district. 

```{r echo=FALSE}
# create new variable "old mosquitoes"
Dhs_data1$Old_Mosquito <- ifelse(Dhs_data1$Months_Ago > 24, 1, 0)
```


```{r echo=FALSE}
library("dplyr")
# To ensure no issues with haven_labelled variables and calculate district-level altitude:
Dhs_data1 <- Dhs_data1 %>%
  mutate(
    Distr = as.character(District),                 
    `Cl_Altitude_ M` = as.numeric(`Cl_Altitude_ M`) 
  ) %>%
  group_by(Distr) %>%
  mutate(
    Average_District_Altitude = mean(`Cl_Altitude_ M`) 
  ) %>%
  ungroup()

Dhs_data1
```




## Data visualizations: 

Produce visualization of your choice. At least each of these
- Bar plot
```{r echo=FALSE}
# your code
# Get the frequency count of each level of Region
# Calculate the count of households by region
count <- table(Dhs_data1$Region)
# Create the bar plot
barplot(count, 
        names.arg = names(count),        
        col = "lightblue",               
        main = "Distribution of Households by Region",  
        xlab = "Region",                 
        ylab = "Frequency",                  
        cex.names = 0.6)                 
                 
```

```{r}
library(dplyr)

# Summarize data by Region and Blood Smear Test
DHS_summary <- Dhs_data1 %>%
  group_by(Region, Educ_level) %>%
  summarize(Count = n(), .groups = "drop") # Use n() for row count if 'Count' column is absent


```


```{r}
library(ggplot2)

# Barplot
ggplot(DHS_summary, aes(x = Region, y = Count, fill = Educ_level)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Blood Smear Test Results by Region",
    x = "Region",
    y = "Count",
    fill = "Educ_level"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5), # Center the title
    legend.position = "top"                # Place legend at the top
  )

```








- Pie plot
```{r echo=FALSE}
# your code
# Create frequency table for regions
count <- table(Dhs_data1$Educ_level) 

percentages <- round(100 * count / sum(count), 1)  


labels <- paste(names(count), "(", percentages, "%)", sep = "")

# Create pie chart with percentages in labels
pie(count, 
    labels = labels,  
    col = rainbow(length(count)),  
    main = "Education Level  Distribution",  
    cex = 0.6)  

```

- Histogram
```{r echo=FALSE}
# your code

# histogram on continous data

Dhs_data1$Age <- gsub("[^0-9.]", "", Dhs_data1$Age)  
age_numeric <- as.numeric(Dhs_data1$Age)            
age_numeric <- na.omit(age_numeric)                 

# Create the histogram
hist(age_numeric, 
     freq = FALSE,        
     main = "Histogram of Age",  
     xlab = "Age",        
     col = "darkblue",    
     border = "black")    


```

- Boxplot
```{r echo=FALSE}
boxplot(Dhs_data1$`Cl_Altitude_ M`, data = data,
        main = "Boxplot of Values by Cluster Altitude",  # Title
        xlab = "Group",  # X-axis label
        ylab = "Values",  # Y-axis label
        col = c("skyblue", "pink", "lightgreen"))  # Colors for boxes
```




```{r echo=FALSE}
# Assuming 'Cluster_Altitude' is the column containing altitude data
Q1 <- quantile(Dhs_data1$`Cl_Altitude_ M`, 0.25, na.rm = TRUE) # 1st quartile
Q3 <- quantile(Dhs_data1$`Cl_Altitude_ M`, 0.75, na.rm = TRUE) # 3rd quartile
IQR <- Q3 - Q1                                                # Interquartile range

# Define outlier boundaries
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

```



```{r echo=FALSE}
Dhs_data_filtered <- Dhs_data1[
    Dhs_data1$`Cl_Altitude_ M` >= lower_bound & 
   Dhs_data1$`Cl_Altitude_ M`<= upper_bound, 
]

```



```{r echo=FALSE}

# After filtering
boxplot(Dhs_data_filtered$`Cl_Altitude_ M`, 
        main = "Boxplot of Values by Cluster Altitude\nAfter Removing Outliers",
        col="skyblue")

```







## Statistical analysis

### Descriptive statistics

1. Use Variable "HML33" to filter out people who had Malaria measurement. 
```{r echo=FALSE}
# your code
tested_positive <- Dhs_data1[Dhs_data1$Malaria_measurement == "Measured",]
```

2. Calculate Malaria [Prevalence](https://dhsprogram.com/data/Guide-to-DHS-Statistics/Prevalence_of_Malaria_in_Children.htm) for both "Blood Smear" and "Rapid Test"
```{r echo=FALSE}
# your code
# Calculate prevalence for Blood Smear
blood_smear_total<-length(Dhs_data1$Blood_smear_test)
blood_smear_positive <- sum(Dhs_data1$Blood_smear_test == "Positive")  
blood_smear_prevalence <- (blood_smear_positive / blood_smear_total) * 100  

# Calculate prevalence for Rapid Test
rapid_test_total<-length(Dhs_data1$`Malaria _rapid_test`)
rapid_test_positive <- sum(Dhs_data1$`Malaria _rapid_test` == "Positive")  
rapid_test_prevalence <- (rapid_test_positive / rapid_test_total) * 100  


cat("Blood Smear Prevalence (%):", blood_smear_prevalence, "\n")
cat("Rapid Test Prevalence (%):", rapid_test_prevalence, "\n")

```

3. Aggregate Prevalence at district Level (Blood smear test)
```{r echo=FALSE}
# your code
library(dplyr)

District_prevelence<-Dhs_data1%>% 
  group_by(District) %>% 
  summarize(
    total=n(),
    positive_cases = sum(Blood_smear_test == "Positive", na.rm = TRUE),
    prevalence = round((positive_cases/total)*100,3)
  )
District_prevelence
```

Prevalence at district Level (Malaria rapid test)
```{r echo=FALSE}
library(dplyr)

District_prevelence<-Dhs_data1%>% 
  group_by(District) %>% 
  summarize(
    total=n(),
    positive_cases = sum(`Malaria _rapid_test` == "Positive", na.rm = TRUE),
    prevalence = round((positive_cases/total)*100,3)
  )
District_prevelence
```



### Analytical Analysis
1. Compare the prevalence in both tests and state if they are different.

***Hint:*** Check `?` the documentations for `t.test` and `aov`.
```{r echo=FALSE}
# your code
# Total counts for both tests
# Calculate prevalence rates (as proportions)
blood_smear_rate <- Dhs_data1$Blood_smear_test == "Positive"
rapid_test_rate <- Dhs_data1$`Malaria _rapid_test` == "Positive"


# Perform an independent t-test
t_test <- t.test(blood_smear_rate, rapid_test_rate)

# Print Results
cat("T-Test Results:\n")
print(t_test)

# Interpretation
if (t_test$p.value < 0.05) {
  cat("The mean prevalence rates are significantly different (p-value:", t_test$p.value, ").\n")
} else {
  cat("The mean prevalence rates are not significantly different (p-value:", t_test$p.value, ").\n")
}

```
The mean prevalence rate for rapid tests is significantly higher than that for blood smear tests. The difference is statistically significant, with a very small p-value (5.579×10−135.579 \times 10^{-13}5.579×10−13), meaning this result is unlikely to have occurred by chance.The observed difference in means (−0.00206-0.00206−0.00206) suggests that the blood smear test consistently detects malaria at a lower rate than the rapid


### Bonus

2. Using a statistical model of your choice, determine if there is a relationship between malaria prevalence in a district and its average altitude. 
```{r echo=FALSE}
# your code
# Aggregate malaria prevalence and altitude by district
District_prevalence <- Dhs_data1 %>%
  group_by(District) %>%
  summarise(
    Malaria_Prevalence = mean(Blood_smear_test == "Positive", na.rm = TRUE) * 100,
    Average_Altitude = mean(`Cl_Altitude_ M`, na.rm = TRUE)
  )
District_prevalence
```





```{r echo=FALSE}
# Fit the linear regression model
model <- lm(Malaria_Prevalence ~ Average_Altitude, data = District_prevalence)


summary(model)
```


```{r echo=FALSE}
# Scatterplot with regression line
plot(District_prevalence$Average_Altitude, District_prevalence$Malaria_Prevalence,
     xlab = "Average Altitude (meters)", ylab = "Malaria Prevalence (%)",
     main = "Malaria Prevalence vs. Average Altitude")
abline(model, col = "red", lwd = 2)  # Add regression line

```
The regression results indicate no significant relationship between altitude and malaria prevalence at the district level. 
While the coefficient for altitude is negative, it is not statistically significant, and the model explains very little of the variability in malaria prevalence 












